Study of Intra- and Inter-Job Interference with Different Job Allocations

### 2016-07-08 app_study.tex
pass on Introduction, remove some citations

A parallel application usually conforms to a combination of several basic communication patterns \cite{roth}. At its different execution phases, the application's communication behavior may follow different basic patterns respectively. There are many profiling tools available to capture information regarding communication patterns of parallel applications~\cite{tau,mpip,scala,sst,oxbow}. They can provide information such as the percentage of different MPI operations of the applications, communication topology, the amount of data transferred between processes.

There are many applications running on HPC systems, such as Mira \cite{bgq} and Titan \cite{titan}, that are communication intensive and conform to various communication patterns. In this work, we select three representative applications from the DOE Design Forward Project. Each application exhibits a distinctive communication pattern that is commonly seen in HPC applications. We believe that the communication patterns of these applications are representative of a wide array of applications running on leadership-class machines. Specifically, we study the Algebraic MultiGrid Solver (AMG), Geometric MultiGrid (MultiGrid) and CrystalRouter MiniApps. The communication pattern figures we present in Section~\ref{sec:amg}, \ref{sec:crystalrouter}, and \ref{sec:multigrid} are generated with the IPM~\cite{ipm} data gathered from publicly available traces~\cite{designforwardwebpage}.

@@

\begin{figure*}[htp]
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[height=1.5in]{figs/appstudy/amg/amg_pip}
        \caption{AMG}
        \label{fig:amg-communication-topology}
    \end{subfigure}
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[height=1.5in]{figs/appstudy/cr/cr_pip}
        \caption{CrystalRouter}
        \label{fig:cr-communication-topology}
    \end{subfigure}
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[height=1.5in]{figs/appstudy/mg/mg_pip}
        \caption{MultiGrid}
        \label{fig:mg-communication-topology}
    \end{subfigure}
    \caption{Applications rank-to-rank communication topology graph.}
\end{figure*}

>>>>

A parallel application usually conforms to a combination of several basic communication patterns~\cite{roth}. 
At its different execution phases, 
the application's communication behavior may follow certain basic patterns respectively. 
There are many profiling tools available to capture 
information regarding communication patterns of parallel applications~\cite{tau,mpip,sst,oxbow}. 
They can provide information such as the percentage of different MPI operations, 
communication topology, the amount of data transferred between processes, etc.

There are many applications running on HPC systems such as Mira~\cite{bgq} and Titan~\cite{titan}. 
These applications are communication intensive and conform to various communication patterns. 
In this work, we select three representative applications from the DOE Design Forward Project. 
Each application exhibits a distinctive communication pattern that is commonly seen in HPC applications. 
We believe that the communication patterns of these applications 
are representative of a wide array of applications running on leadership-class machines. 
Specifically, we study the Algebraic MultiGrid Solver (AMG), 
Geometric MultiGrid (MultiGrid) and CrystalRouter MiniApps. 
The communication matrix presented in Figure~\ref{fig:apps_communication_matrix}
are generated with the IPM~\cite{ipm} data gathered from publicly available traces~\cite{designforwardwebpage}.

@@

\begin{figure*}[htp]
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[height=1.5in]{figs/appstudy/amg/amg_pip}
        \caption{AMG}
        \label{fig:amg-communication-topology}
    \end{subfigure}
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[height=1.5in]{figs/appstudy/cr/cr_pip}
        \caption{CrystalRouter}
        \label{fig:cr-communication-topology}
    \end{subfigure}
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[height=1.5in]{figs/appstudy/mg/mg_pip}
        \caption{MultiGrid}
        \label{fig:mg-communication-topology}
    \end{subfigure}
    \caption{Applications rank-to-rank communication topology graph.}
    \label{fig:apps_communication_matrix}
\end{figure*}

### 2016-07-08 discussion.tex
pass on Introduction, remove some citations

The advantage of making allocation with consideration about job's communication pattern is obvious. First, compared with contiguous allocation policy, the new scheduler outlined above is more flexible, it doesn't require the system to provide a big contiguous partition to accommodate the whole application, instead it only needs to provide a small set of compact nodes that is sufficient for all the "local community" in the application. On the other hand, the new scheduler outlined above won't destroy the locality of application's communication pattern. Indeed, the small compact node set (allocation unit) provided for the application's  ``local communities" can preserve the communication locality in the maximum extent. The design of such a smart scheduler is part of our future work.

>>>>

The advantage of making allocation with consideration about job's communication pattern is obvious. First, compared with contiguous allocation policy, the new scheduler outlined above is more flexible, without requiring the system to provide a big contiguous partition to accommodate the whole application, it only needs to provide a small set of compact nodes that is sufficient for all the "local community" in the application. On the other hand, the new scheduler outlined above won't destroy the locality of application's communication pattern. Indeed, the small compact node set (allocation unit) provided for the application's  ``local communities" can preserve the communication locality in the maximum extent. The design of such a smart scheduler is part of our future work.

### 2016-07-08 interference_study.tex
pass on Introduction, remove some citations

Lots of research studies have designed fancy allocation algorithms to provide application with the most compact allocation\cite{leung,LO}. However, as shown in our experiments, providing cubical allocation without considering application's communication pattern won't guarantee the best performance for every application. Compact allocation should be provided to application with global data transfer, such as those conforms to ``Many-to-Many" communication pattern. Application with dominant communication pattern like ``Nearest Neighbor" won't fully utilize the compact allocation and could get even better performance on a relaxed allocation shape.

@@

The ``Cube" mapping benefits MultiGrid's ``Many-to-Many" communication. However, due to the small amount of data transfer among ranks in MultiGrid, ``Cube" mapping doesn't exhibit a drastic advantage. ``Random" mapping neither cause too much degradation for the same reason, shown in Figure \ref{fig:diffmap-mg}.

>>>>

Lots of research studies have designed complex allocation algorithms to provide application with the most compact allocation\cite{leung,LO}. However, as shown in our experiments, providing cubical allocation without considering application's communication pattern can not guarantee the best performance for every application. Compact allocation should be provided to application with global data transfer, such as those conforms to ``Many-to-Many" communication pattern. Application with dominant communication pattern like ``Nearest Neighbor" won't fully utilize the compact allocation and could get even better performance on a relaxed allocation shape.

@@

The ``Cube" mapping benefits MultiGrid's ``Many-to-Many" communication. However, due to the small amount of data transfer among ranks in MultiGrid, ``Cube" mapping fails to exhibit a drastic advantage. ``Random" mapping neither cause too much degradation for the same reason, shown in Figure \ref{fig:diffmap-mg}.

### 2016-07-08 intro.tex
pass on Introduction, remove some citations

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.2\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/goodallocation}
        \caption{Contiguous}
        \label{fig:overview_sub1}
    \end{subfigure}%
    \hspace{1em}%
    \begin{subfigure}[t]{0.2\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/badallocation}
        \caption{Non-contiguous}
        \label{fig:overview_sub2}
    \end{subfigure}%
   \caption{Multiple jobs running concurrently with different allocations. Each job is represented by a specific color. a) shows the effect of contiguous allocation, which reduce the inter-job interference. b) shows non-contiguous allocation, which may introduce both intra and inter-job interferences. }
   \label{fig:overview}
\end{figure}

When submitting jobs to HPC systems, users request resources in terms of node counts and expected runtime. The batch scheduler is then responsible for dispatching the submitted jobs to the system. Typically, multiple jobs run concurrently, resulting in the shared use of resources, particularly network and storage. A prominent problem with this approach is network contention among concurrently running jobs. If the network resources of concurrently running jobs aren't isolated, the resulting sharing can cause communication variability that results in performance degradation for some or all such jobs \cite{abhinav-sc13}. This delay can propagate into the queueing time of the following submitted jobs, thus leading to low system throughput and utilization \cite{jose-ipdps15}. This adverse effect of network sharing can be mitigated by providing jobs with isolated allocations and exclusive network resources.

On the widely used torus-connected HPC systems \cite{bgq,tofu,titan}, two allocation strategies are commonly used. The \emph{contiguous allocation} strategy assigns to each job a compact and contiguous set of computing nodes, as shown in Figure \ref{fig:overview_sub1}. The partition-based allocation approach used in Blue Gene series systems is an example of such a strategy \cite{bgloverview}. The contiguous strategy favors application performance through isolated networking within a partition and the locality that implies. However, this strategy can cause both internal fragmentation (when more nodes are allocated to a job than it requests) and external fragmentation (when sufficient nodes are available for a request, but they can not be allocated contiguously), therefore leading to lower system utilization than is otherwise possible. On the other side of the coin, the \emph{non-contiguous allocation} strategy, used by the Cray XT/XE series \cite{carl-cug}, assigns free nodes to jobs regardless of contiguity, though of course efforts are made to maximize locality. Figure~\ref{fig:overview_sub2} shows an example non-contiguous allocation. While eliminating internal and external fragmentation as seen in contiguous allocation systems, in return non-contiguous allocations introduce contention between jobs due to the interleaving of job nodes. The non-contiguous node allocation can significantly reduce job performance, especially for communication-intensive ones \cite{abhinav-sc13}.

We envision that future HPC schedulers will adopt a flexible job allocation mechanism which combines the best of both contiguous and non-contiguous allocation strategies. Such a flexible mechanism would take shared resource needs of jobs into account when making allocation decisions (e.g., network). With knowledge and analysis of job communication patterns, it can be identified which jobs require network isolation and locality, and to what degree. Then, rather than allocating each job in a ``know-nothing'' manner, one may specialize allocation so that, for example, only the jobs with stringent network needs are given compact, isolated allocations, resulting in maximized utilization and minimized perceivable resource contention effects.

In this work, we focus on an in-depth analysis of intra- and inter-job communication interference with different job allocation strategies on torus-connected HPC systems. Torus-based networks are used on six of the top 10 supercomputers on the June 2015 Top500 lists~\cite{top500}. The current generation of IBM Blue Gene/Q (BG/Q) supercomputers, such as Mira at Argonne Nation Laboratory and Sequoia at Lawrence Livermore National Laboratory, has its nodes connected in a 5D torus network \cite{bgq}. The K computer from Japan uses the ``Tofu'' system, which is a 6D mesh/torus topology \cite{tofu}. Titan, a Cray XK7 supercomputer located at the Oak Ridge Leadership Computing Facility (OLCF), has nodes connected in a 3D torus within the compute partition \cite{titan}. Although our analyses are based on torus networks, the ideas conveyed in this work is applicable to networks with different topologies.

We selected three signature applications from the DOE Design Forward Project~\cite{designforwardwebpage} as examples to conduct detailed study about their communication patterns. We use a sophisticated simulation toolkit named CODES (standing for Co-Design of Multi-layer Exascale Storage Architectures)~\cite{Jason-2011} as a research vehicle to evaluate the performance of these applications with various allocations in a controlled environment. We then analyze the intra- and inter-job interference by simulating these applications running exclusively and concurrently with different allocations. We believe the insights presented in this work can be very useful for the design of future HPC batch job schedulers and resource managers.

%\NOTE{JJ removed the list of contributions, because they are exactly the same as the preceding paragraph.}
\begin{comment}
The observations of this work are the following:

\begin{itemize}
    \item We study the communication behaviors and identify the communication patterns of three representative HPC applications.
    \item We conduct detailed analysis about the intra- and inter-job interference by running three applications with different allocation strategies. And we find the relation between application's communication pattern and the cause of interference.
    \item We show a novel and flexible way to make allocation with awareness of job's communication pattern, which  is every effective to alleviate the interference between concurrently running jobs. \TODO{do we actually do this, or provide the underpinnings for such a method?}
\end{itemize}

We believe the finds and insights presented in this work would be very useful for the design of future HPC batch job scheduler and resource management module.

\end{comment}

The rest of this paper is organized as follows. Section~\ref{sec:application study} describes the three representative applications from DOE Design Forward Project for our study. Section~\ref{sec:codes} talks about the use of CODES as research vehicle for our work. Section~\ref{sec:config study} shows the performance analysis of three applications on different torus networks. Section~\ref{sec:interference} provides detailed analysis about the intra- and inter-job interference between three applications on torus network with different allocation strategies. Section~\ref{sec:discussion} introduces a path toward communication-pattern aware allocation strategies, given the results of our analysis. Section~\ref{sec:related_work} discusses related work. Finally, the conclusion is presented in Section~\ref{sec:conclusion}.

>>>>

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/goodallocation}
        \caption{Contiguous}
        \label{fig:overview_sub1}
    \end{subfigure}%
    \hspace{1em}%
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/badallocation}
        \caption{Non-contiguous}
        \label{fig:overview_sub2}
    \end{subfigure}%
   \caption{Multiple jobs running concurrently with different allocations. 
   Each job is represented by a specific color. 
   a) shows the effect of contiguous allocation, 
   which could possibly reduce the inter-job interference. 
   b) shows non-contiguous allocation, 
   which may introduce both intra and inter-job interference. 
   }
   \label{fig:overview}
\end{figure}

When submitting jobs to HPC systems, 
users request system resources by specifying the number of compute nodes and expected runtime. 
The batch scheduler is then responsible for dispatching the submitted jobs to the system. 
Typically, there are multiple jobs running concurrently on the system,
resulting in the shared use of resources, particularly network links. 
A prominent problem with this network sharing is the contention over
the network resource among those concurrently running jobs. 
The network contention can cause communication variability and 
performance degradation to those jobs~\cite{abhinav-sc13}. 
This performance degradation can propagate into the queueing time of the following submitted jobs, 
thus leading to low system throughput and utilization~\cite{jose-ipdps15}. 
This adverse effect of network sharing can be mitigated by 
providing jobs with isolated allocations and exclusive network resources.

On the widely used torus-connected HPC systems~\cite{bgq,tofu,titan}, 
two job placement policies are commonly used. 
In \emph{contiguous placement} policy, 
each job gets a compact and contiguous set of computing nodes, 
as shown in Figure~\ref{fig:overview_sub1}. 
The partition-based placement adopted on the Blue Gene 
series systems is an example of contiguous placement policy~\cite{bgloverview}. 
The contiguous placement favors application performance 
through exclusive networking within a compact partition 
and the locality that implies. 
However, contiguous placement can cause both internal fragmentation 
(when more nodes are allocated to a job than it requests) 
and external fragmentation 
(when sufficient nodes are available for a request, but they can not be allocated contiguously), 
therefore leading to lower system utilization than is otherwise possible. 
On the other hand, the \emph{non-contiguous placement} policy, 
adopted by the Cray XT/XE series~\cite{carl-cug}, 
assigns free nodes to jobs regardless of contiguity, 
though of course efforts are made to maximize locality. 
Figure~\ref{fig:overview_sub2} shows the effect of non-contiguous placement. 
While eliminating internal and external fragmentation as seen in contiguous allocation systems, 
in return non-contiguous allocations introduce contention between jobs due to the interleaving of job nodes. 
The non-contiguous job placement can significantly reduce job performance, 
especially for communication-intensive ones~\cite{abhinav-sc13}.

We envision that future HPC systems will be equipped with 
a flexible job placement mechanism which combines the 
best of both contiguous and non-contiguous placement. 
Such a flexible job placement mechanism should take 
shared resource needs (e.g., network resources) of jobs into account 
when making scheduling and allocation decisions. 
With the knowledge and analysis of job communication patterns, 
it can be identified which jobs require exclusive network and allocation locality, and to what degree. 
Then, rather than allocating each job in a ``know-nothing'' manner, 
one may specialize allocation so that, for example, 
only the jobs with stringent network needs are given compact, isolated allocations, 
resulting in maximized utilization and minimized perceivable resource contention effects.

In this work, we focus on an in-depth analysis of intra- and inter-job 
communication interference with different job placement on torus-connected HPC systems. 
Torus-based networks are used on six of the top 10 supercomputers 
on the June 2015 Top500 lists~\cite{top500}. 
The current generation of IBM Blue Gene/Q (BG/Q) supercomputers, 
such as Mira at Argonne Nation Laboratory and 
Sequoia at Lawrence Livermore National Laboratory, 
have their compute nodes connected by a 5D torus network~\cite{bgq}. 
The K computer from Japan uses the ``Tofu'' interconnected system, 
which has a 6D mesh/torus topology~\cite{tofu}. 
Titan, a Cray XK7 supercomputer located at the Oak Ridge Leadership Computing Facility (OLCF), 
has nodes connected in a 3D torus within the compute partition~\cite{titan}. 
Although our analyses are based on torus networks, 
the ideas conveyed in this work is applicable to networks with different topologies.

We selected three signature applications from 
the DOE Design Forward Project~\cite{designforwardwebpage} as examples 
to conduct detailed study about the communication pattern of parallel applications. 
We use a sophisticated simulation toolkit named CODES 
(standing for Co-Design of Multi-layer Exascale Storage Architectures)~\cite{Jason-2011} 
as a research vehicle to evaluate the performance of these applications 
with various allocations in a controlled environment. 
We then analyze the intra- and inter-job interference 
by simulating these applications running concurrently with different allocations. 
We believe the insights presented in this work can be very useful 
for the design of future HPC batch schedulers and resource managers.

The rest of this paper is organized as follows. 
Section~\ref{sec:application study} describes the three representative applications 
from the DOE Design Forward Project for our study. 
Section~\ref{sec:codes} talks about the use of CODES as research vehicle for our work. 
%Section~\ref{sec:config study} shows the performance analysis of three applications on different torus networks. 
Section~\ref{sec:interference} provides detailed analysis about the intra- and inter-job interference 
among three applications on torus network with different allocations. 
Section~\ref{sec:discussion} introduces a path toward communication-pattern aware allocation strategies, 
given the results of our analysis. 
Section~\ref{sec:related_work} discusses related work. 
Finally, the conclusion is presented in Section~\ref{sec:conclusion}.

### 2016-07-08 related_work.tex
pass on Introduction, remove some citations

There are many tools available for system monitoring and application profiling. Tools like TAU (Tuning and Analysis) \cite{tau}, SCALASCA \cite{scala} and mpiP \cite{mpip} can capture application's runtime information about application's communication and computation in event traces. However, recognizing the patterns about application's communication from those traces requires substantial effort. There are a number of studies on the recognition and characterization of parallel application communication patterns. Oak Ridge National Laboratory has its ongoing project about developing a tool set named Oxbow, which can characterize the computation and communication behavior of scientific applications and benchmarks \cite{oxbow}. In a recent work \cite{roth}, the authors demonstrate a new approach to automatically characterizing parallel applications' communication behaviors.

>>>>

There are many tools available for system monitoring and application profiling. Tools like TAU (Tuning and Analysis)~\cite{tau}, 
and mpiP~\cite{mpip} can capture application's runtime information about application's communication and computation in event traces. However, recognizing the patterns about application's communication from those traces requires substantial effort. There are a number of studies on the recognition and characterization of parallel application communication patterns. Oak Ridge National Laboratory has its ongoing project about developing a tool set named Oxbow, which can characterize the computation and communication behavior of scientific applications and benchmarks~\cite{oxbow}. In a recent work~\cite{roth}, the authors demonstrate a new approach to automatically characterizing parallel applications' communication behaviors.
